---
title: "Analysis of New York City Shooting Incidents"
date: "June 2, 2021"
author: "Elise Geissler-Huesgen"
subtitle: "Is year (i.e. time) a good predictor for shooting incidents?"
output:
  pdf_document: default
  word_document: default
  html_document: default
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Background

Gun violence continues to be a major problem in New York City, which saw
1,942 shootings in the year 2020 alone. If data of past shootings could predict
when and where shootings are more likely to occur, then finite resources 
earmarked for gun violence prevention could be distributed in a more targeted 
manner. 

The dataset utilized for this project represents a list of every shooting 
incident which has taken place in New York City between the years of 2006 
and 2020. Each row includes information on timing and location of the 
incident, as well as demographic data on the victim and the perpetrator. It 
is an open-access dataset which is updated and published quarterly by the 
New York City Police Department. 

The following analysis asks the question, whether time is a good predictor for
the number of shooting incidents that are going to occur. In other words, do 
shooting incidents go up or down in a predictable manner through the years? 


#  Setting up

Loading all necessary libraries

```{r message = FALSE, warning = FALSE, label = "libraries"}
library(tidyverse) #data wrangling
library (lubridate) #dates and times
library(ggthemes) #ggplot themes
library(broom) #for tidy() and map()

#libraries for NYC map with bubble graph:
library(httr) #for GET()
library(rgdal) #for readOGR()
library(geosphere)#for controid()
```


And loading the data:


```{r, label = "data link"}

link <- "https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD"
raw_shootingData <- read.csv(link)

head(as_tibble(raw_shootingData)) #taking a look at the data

```

Selecting columns:

```{r message = FALSE, warning = FALSE, label = "column select"}
shooting_data <- raw_shootingData %>% 
                select(-c(INCIDENT_KEY, OCCUR_TIME, 
                          PRECINCT:LOCATION_DESC,PERP_RACE,
                          VIC_RACE, X_COORD_CD:Lon_Lat))
```

Renaming some of the longer column names:


```{r}
names(shooting_data)[1] <- "DATE" #shortening column name
names(shooting_data)[3] <- "MURDER" #shortening column name
names(shooting_data)[4] <- "PERP_AGE" #shortening column name
names(shooting_data)[6] <- "VIC_AGE" #shortening column name
```


# Data cleaning

## Initial overview of the data


```{r message = FALSE, warning = FALSE, label = "data summary"}
options(knitr.kable.NA = "") #missing variables will be empty cells in output
knitr::kable(summary(shooting_data), caption = "Summary of data") #knitr::kable for pretty output
```

I can already spot some possible trends in data distribution, for example in 
columns concerning age, sex, and location. These trends would be worth 
further exploration in the future. 

Given that there are so many types of data in one dataframe, I want to know 
whether there are any data classes that need to be changed:

```{r message=FALSE, warning = FALSE, label = "data classes"}
knitr::kable(apply(shooting_data, 2, class), caption = "Variable classes",
             col.names = c("Class")) #class by column
```
Everything has been imported as characters. So I will need to change the date
column to class "date". 

## Cleaning up dates

Changing class of date column and adding year, month, and season columns.

```{r message = FALSE, warning = FALSE, label = "date cleaning"}
clean_data <- shooting_data %>%
  mutate(DATE = mdy(DATE)) %>% #change to "date"
  mutate(YEAR = year(DATE)) %>% #adding year column
  mutate(MONTH = month(DATE, label=TRUE)) %>% #adding month column
  mutate(SEASON = hydroTSM::time2season(DATE,out.fmt = "seasons"))%>% #seasons 
  mutate(SEASON = replace(SEASON, SEASON == "autumm", "fall"))%>% #fixing a typo
  select(-DATE) #removing column, it takes up lots of space and I won't need it
knitr::kable(head(clean_data), caption = "First rows of cleaned data") # taking a look @ cleaned data
```

## Missing and unknown variables

I am not sure how missing and unknown variables are coded in this dataset. 
Therefore, I am initially looking through all unique variables per column.


```{r message = FALSE, warning = FALSE, label = "unique variables"}
apply(clean_data, 2, unique)  #prints unique variables for each column
```

The dataset includes "U", UNKNOWN", and empty variables. 

I will initially tackle the empty variables. Instead of deleting them, 
I will code them as "NAs", as those are easy to find and easy to delete. 
Empty values are limited to columns 3 and 4.

```{r message = FALSE, warning = FALSE, label = "empty na"}
clean_data[,3:4][clean_data[,3:4] == ""] <- NA #changing "" to "NA" in cols 3, 4
```

## "NA" variables

To see how many NAs there are:

```{r message = FALSE, warning = FALSE, label = "sum na"}
colSums(is.na(clean_data[,3:4])) %>% as.data.frame() 
```

There are a lot of NAs in the dataframe. As their presence will not affect any
of the analyses in this document and deleting them would greatly reduce the data,
I will not delete the affected rows. 

## "UNKOWN"/"U" variables

There are also rows which include "U"/"UNKNOWN", which should not be conflated 
with the NA rows. For example, "unknowns" are likely the result of an 
unidentified or anonymized perpetrator or victim, whereas NAs could simply be 
the result of someone not filling in the variable. However, I am curious 
whether NAs and "unknowns" tend to show up on the same row:

```{r message = FALSE, warning = FALSE, label = "rows na unknowns"}
na_unknowns <- clean_data %>% 
  mutate (NAs = ifelse(rowSums(is.na(.[,3:4])) > 0,"Present", "Absent")) %>%
  mutate (Unknowns = ifelse(apply(clean_data, 1, function(x) 
                            any (x %in% c("UNKNOWN", "U"))),"Present", "Absent")) %>%
  group_by(NAs, Unknowns) %>%
  summarise(Count = n()) %>%
  mutate(Ratio_NaUnk = paste0(round(100*Count/sum(Count),0), "%"))
knitr::kable(na_unknowns, caption = "Rows with NAs and unknowns")
```

The table shows the percentage of rows that have an NA value (= present) that also have 
an "unknown" and the percentage of those that have an NA, but do not have an unknown. 
The same analysis was carried out for rows that do not have NAs present (= absent). 

Only 1% of rows which have an "unknown" or "u" variable, also have an NA. This 
result confirms my suspicion that I should leave these two variables as 
separate variables. 


# Visualizing some variables

## Shooting incidents by year and victim survival

I will start with the year of the shooting and whether the victim has survived
the shooting. 

First, I need the number of shooting incidents per year and whether the victim
survived: 

```{r message = FALSE, warning = FALSE}
survivalYear <- clean_data %>% 
  group_by(YEAR,MURDER) %>% #grouping by year and survival
  summarise(VictimCount = n()) %>% #counting rows for each group
  ungroup()%>%
  group_by(YEAR) %>% #grouping by just year
  mutate(Ratio = paste0(round(100*VictimCount/sum(VictimCount),0), "%")) #% survival vs. killed
knitr::kable(head(survivalYear), caption= "First rows of survivors vs. murder victims") 
```

And the graph: 

```{r message = FALSE, warning = FALSE}
ggplot(survivalYear, aes(y=VictimCount, x=factor(YEAR), fill = MURDER))+
  geom_bar(stat="identity")  +
  labs(x = "Year", y = "Shooting incidents", fill="Victims",
       title = "Shooting Incidents in New York City, 2006-2020",
       subtitle = "Comparing survivors and non-survivors") +
  geom_text(aes(label = Ratio, vjust = 1.4), size = 4) + 
  theme_minimal()  + 
  theme(legend.position = "bottom") +
  scale_fill_brewer (palette="Accent", labels = c("Survived", "Killed")) 
```


## Shooting incidents by season

Counting the number of incidents per season, followed by a bar plot:

```{r message = FALSE, warning = FALSE}
seasonShootings <- clean_data %>% 
  group_by(SEASON) %>% 
  summarise(BySeason = n()) #number of shooting incidents by season

ggplot(seasonShootings, aes(y=BySeason, x=factor(SEASON)))+
  geom_bar(stat="identity", fill='palegreen3')  +
  labs(x = "Season", y = "Shooting incidents",
       title = "Shooting Incidents in New York City by season",
       caption = "Data collection: 2006-2020") +
  theme_minimal()  
```

## Month of shooting incident and victim survival

Looking at the combination of month of shooting incidents and whether the
victim survived:

```{r message = FALSE, warning = FALSE}
survivalMonth <- clean_data %>% 
  group_by(MONTH,MURDER) %>% 
  summarise(SuMo = n()) %>% #number of shootings per month and per outcome
  ungroup()%>%
  group_by(MONTH) %>%
  mutate(Ratio = paste0(round(100*SuMo/sum(SuMo),0), "%")) #ratio of survival vs. murder per month

ggplot(survivalMonth, aes(y=SuMo, x=factor(MONTH), fill = MURDER))+
  geom_bar(stat="identity")  +
  labs(x = "Month", y = "Shooting incidents", fill="Victims",
       title = "Shooting Incidents in New York City by Month",
       subtitle = "Comparing survivors and non-survivors",
       caption = "Data collection: 2006-2020") +
  geom_text(aes(label = Ratio, vjust = 1.3), size = 4) + 
  theme_minimal()  +
  theme(legend.position = "bottom") +
  scale_fill_brewer (palette="Accent", labels = c("Survived", "Killed")) 
```

## Male vs. female victims 

Shooting incidents through time, comparing male and female victims

```{r message = FALSE, warning = FALSE}
shootingsFeMa <- clean_data %>% 
  group_by(YEAR,VIC_SEX) %>% 
  summarise(FeMa = n()) #shootings per year, per victim sex

ggplot(data = shootingsFeMa, aes(x=factor(YEAR), 
                                 y = FeMa, group = VIC_SEX, col = VIC_SEX)) +
      geom_line(size=1) +
      labs(x = "Year", y = "Shooting victims", col = "Victims",
           title = "Comparing Male and Female Shooting Victims by Year") +
      theme_minimal()  +
  theme(legend.position = "bottom") +
  scale_colour_brewer (palette="Accent", labels = c("Female", "Male", "Unknown"))
```

## Comparing shooting incidents across boroughs

Shooting incidents per year, comparing different boroughs

```{r message = FALSE, warning = FALSE}
timeBoroughs <- clean_data %>% 
  group_by(YEAR,BORO) %>% 
  summarise(TimeBo = n()) #shootings per year per borough

ggplot(data = timeBoroughs, aes(x=factor(YEAR), 
                                 y = TimeBo, group = BORO, col = BORO)) +
  geom_line(size=1) +
  labs(x = "Year", y = "Shooting victims", col = "Borough",
       title = "Shooting Incidents in New York City",
       subtitle = "Comparing shooting incidents across boroughs") +
  theme_minimal()  +
  theme(legend.position = "bottom") +
  scale_colour_brewer (palette="Accent")
```

## Shooting incident map

And finally, a map comparing the total shooting incidents for each New York
City borough, between 2006-2020.The following graph will summarize the overall 
number of shooting incidents for each borough throughout the time frame of the 
data collection. The graph above indicates that there may be large differences
and these may be worth further exploration in the future. 

To get shape data for each borough:

```{r message = FALSE, warning = FALSE}
shapeLink <- 'http://data.beta.nyc//dataset/0ff93d2d-90ba-457c-9f7e-39e47bf2ac5f/resource/35dd04fb-81b3-479b-a074-a27a37888ce7/download/d085e2f8d0b54d4590b1e7d1f35594c1pediacitiesnycneighborhoods.geojson'
nyc_shape_data <- GET(shapeLink)
#Making loaded shape data suitable for R:
nyc_shapes <- readOGR(content(nyc_shape_data,'text'), 'OGRGeoJSON', verbose = F)
```

Turning the shape data into a data frame:

```{r message = FALSE, warning = FALSE}
nyc_shapes_df <- tidy(nyc_shapes, region = "borough") #convert to data frame
knitr::kable(head(nyc_shapes_df), caption = "First rows of NYC shape data as data frame")
```

Calculating centroids and merging those with incident numbers, so the map can be
combined with a bubble plot. 

```{r message = FALSE, warning = FALSE}
borough_data <- nyc_shapes_df %>%
  group_by (id) %>%
  mutate (Centroid = centroid(as.matrix(cbind(long, lat)))) %>% #centroid calculation
  ungroup()%>%
  group_by (id, Centroid) %>%
  summarise(CasesByBoro = n()) #cases by borough and centroid
knitr::kable(borough_data, caption = "Centroids of boroughs and shooting incidents")
```

And finally the plot:

```{r message = FALSE, warning = FALSE}
ggplot() + 
  geom_polygon(data=nyc_shapes_df, aes(x=long, y=lat, group=group,  fill = id))+
  labs(fill = "Borough", title = "Shooting Incidents in New York City, 2006-2020") +
  geom_point(data = borough_data, aes(x = Centroid[,1], y = Centroid[,2], 
                                      size = CasesByBoro), color = "grey", 
             alpha = 0.7, show.legend = FALSE) +
  scale_size(range = c(15, 24)) +
  geom_text(data = borough_data, aes(x = Centroid[,1], y = Centroid[,2], 
                                     label = borough_data$CasesByBoro)) +
  theme_map() + theme(legend.position = "bottom")+
  scale_fill_viridis_d ()
```


# Regression analysis

## Does year predict amount of shooting incidents?


First, I will create a data frame containing incident count per year. As NAs are
not present in the year column and incident count is not affected by NAs,  I will 
not have to delete the affected rows. 

```{r message = FALSE, warning = FALSE}
yearly_data <- clean_data %>% 
  group_by(YEAR) %>% 
  summarise(Incident_Count = n()) %>% #incident count by year
  ungroup() 
```

Linear model for incident count by borough

```{r message = FALSE, warning = FALSE}
time_model <- lm(Incident_Count~YEAR, data = yearly_data) #linear regression
summary(time_model) #taking a look at the model
```

**Interpretation of regression results:**

* A p-value of 0.002609 indicates that there is a statistically significant relationship between year and amount of shooting incidents.
HOWEVER:
* Resisuals are not strongly symmetrically distributed, meaning that some of the predicted values are not close to the actual values. Therefore, the model does not fit very well. 
* The model only explains 51% of the variation, as indicated by the R-squared value. 

In summary: While there is a significant relationship between year and shooting incidents, year is not a reliable sole predictor for shooting incidents in the given dataset. 

## Plotting actual vs. predicted data

Merging predictions and actual data:

```{r message = FALSE, warning = FALSE}
combined_real_pred <- yearly_data %>% 
  rename(Actual = Incident_Count) %>%
  mutate(Prediction = predict(time_model)) %>% #adding a prediction column
  pivot_longer(c(Actual, Prediction), names_to = "Nature", values_to = "Victim_count")  #wide to long format
knitr::kable(head(combined_real_pred), caption = "First rows of dataframe with actual and predicted data")
```

Followed by the plot:

```{r}
ggplot (data = combined_real_pred, aes(x = YEAR, y = Victim_count, group = Nature, color = Nature, shape = Nature)) +
  geom_point(size = 3)+
  labs(x = "Year", y = "Shooting Incidents", 
       title = "Shooting Incidents in New York City",
       subtitle = "Comparing actual and predicted values") + 
  theme_minimal()
```

# Conclusion and Bias Identification

The plot demonstrates why year is not a reliable indicator for the amount of 
shooting incidents, even though there is a relationship between these two variables. 
Year is simply just a label for 365 days which encompass countless events, 
developments, and human decisions. Consequently, it makes sense that the data could not be 
reliably predicted by only a year variable. The year 2020 is a good example for this. The 
amount of shooting incidents in 2020 far exceed the predicted number of shooting
incidents. It is unclear whether this spike in shooting incidents is connected 
to the global covid-19 pandemic and whether future years would follow a linear 
trend more reliably. However, one can envision shooting incidents to be affected
by a lot of other factors besides a global pandemic, such as economic and 
political factors, which would also not be year-dependent.

The largest source of bias in this analysis is that it is based on the idea that
time can be a predictor for human decision-making. This bias was mitigated through
a conservative interpretation of the model. Other sources of bias are the selection
of variables which were collected and recorded, and the arbitrary
time frame (2006-2020). Additionally, predictions were made based on the analysis
of the data at hand. However, in order to assess the reliability of the model, it would need
to be also tested on other data. 

It would be interesting to incorporate more variables and to see whether the 
resulting model works for future years. Based on the data at hand, it would be
worthwhile to include variables, such as borough, perpetrator and victim
descriptions into the analysis. 

```{r}
sessionInfo()
```







